---
title: "Tree Methods Lab 1: BART"
author: "Ander Wilson"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, purl=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE)
knit_hooks$set(purl = hook_purl) 
```

## Lab overview

In this lab we conduct an analysis of the NHANES data using two approaches:

- BART with covariates included in the tree model.

We will fit the BART model in the lab. This only takes a minute or so. Some of the post-processing for partial dependence plots is slower. We will skip this. These steps were run in advance and output from these steps will be used in the lab.



## Setup

Load the packages used in this lab.

```{r load packages, message=FALSE}
## load required libraries 
#install.packages("corrplot")
library(corrplot)
#install.packages("ggplot2")
library(ggplot2)
#install.packages("tidyverse")
library(tidyverse)
#install.packages("BART")
library(BART)
```

### Lab-specific package

Uncomment the below and then load `bartmix` library when running for first time

This package has some functions to post-process the MCMC from BART models. You can adapt the code in this package to meet other estimation needs with BART.

```{r load bartmix, message=FALSE}
# install.packages("remotes")
# remotes::install_github("AnderWilson/bartmix")
library(bartmix)
```

## Load and format data

Load in the data and do some data preparation. The packages used here (`randomForests` and `BART`) can handle cateogrical predictors. We therefore do less data preparation because we do not need to make dummy variables.

```{r load data}
## read in data and only consider complete data 
## this drops 327 individuals
## some tree methods handle missing data but we will not deal with that here
nhanes <- na.omit(read.csv(paste0(here::here(),"/Data/studypop.csv")))
```

```{r format outcome}
## our y variable - ln transformed and scaled mean telomere length
## some tree methods make assumptions about the error variance and some do not
## generally safer to transform the dependent variable to reduce heteroskedasticity
lnLTL_z <- scale(log(nhanes$TELOMEAN)) 
```

```{r format exposures}
## exposures matrix
mixture <- with(nhanes, cbind(LBX074LA, LBX099LA, LBX118LA, LBX138LA, LBX153LA, LBX170LA, LBX180LA, LBX187LA, LBX194LA, LBXHXCLA, LBXPCBLA,
                              LBXD03LA, LBXD05LA, LBXD07LA,
                              LBXF03LA, LBXF04LA, LBXF05LA, LBXF08LA)) 
# Most tree models are invariant the transformations of the predictors (they don't make a difference)
# here we transform it to make the plots consistent with other methods so they can be compared
mixture   <- apply(mixture, 2, log)
mixture <- scale(mixture)
colnames(mixture) <- c(paste0("PCB",c(74, 99, 118, 138, 153, 170, 180, 187, 194, 169, 126)), 
                           paste0("Dioxin",1:3), paste0("Furan",1:4)) 
exposure_names <- colnames(mixture)
```

```{r format covariates}
## our X matrix
covariates <- with(nhanes, cbind(age_cent, male, bmi_cat3, edu_cat, race_cat,
                                 LBXWBCSI, LBXLYPCT, LBXMOPCT, 
                                 LBXNEPCT, LBXEOPCT, LBXBAPCT, ln_lbxcot)) 
```





## Fit BART

We fit the BART model with default parameters: 200 trees, 2000 MCMC iteration discarded for burning, 2000 iterations retained for analysis.

```{r fit BART model}
# fit the BART model
set.seed(1000)
fit_bart <- gbart(x.train=cbind(mixture,covariates), 
                  y.train=lnLTL_z,
                  nskip=2000,    # MCMC iterations that are discarded as burning
                  ndpost=2000)   # MCMC iterations after burning that are retained for 

load(paste0(here::here(), "/Supervised/Tree Based Methods/bart_fits_for_lab.rda"))
```

## Fitted values

Fitted values in BART are the estimated mean response at the observed values. Note that this is very different that the random forest plot because the covariates have not been regressed out of the outcome (see other lab document).

```{r fitted values}
plot(fit_bart$yhat.train.mean~lnLTL_z, 
     main="fitted vs observed with BART")
```


## Variable importance

A simple measure of variable importance in BART is how many trees each variable is included in. We get an estimate of this with uncertainty. A variable being included in a lot of trees does not necessarily mean the variable has a large association or an association at all.

BART can be extended to do variable selection (use `sparse=TRUE`). Variable selection will be applied to all predictors including covariates. It may be more appropriate when covariates are regressed out of the outcome. This is likely to lead to more useful measures of variable importance.

The `bartMachines` package has additional variable importance measures.

```{r BART variable importance}
# variable importance for the mixture components with BART
varcount <- data.frame(exposure=exposure_names,
                       mean=fit_bart$varcount.mean[exposure_names],
                       lower=apply(fit_bart$varcount[,exposure_names],2,quantile,0.025),
                       upper=apply(fit_bart$varcount[,exposure_names],2,quantile,0.975))

# visualize the variable importance with BART
ggplot(varcount, aes(x=exposure, y=mean, ymin=lower, ymax=upper)) +
  geom_errorbar(width=0) +
  geom_point() +
  theme_minimal() + 
  ggtitle("Variable Inclusion Count with BART") + 
  xlab("Exposure") + 
  ylab("Count") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```



## Estimation a contrast in exposures

We can estimate an IQR change in furan 1 when all other variables (exposures and covariates) are fixed at the sample means. 

The first step it to setup the data.

```{r BART IQR change at mean of other variables data prep}
# find the means
# note that we really should use mode or something other than mean for factor variables
data_means <- colMeans(cbind(mixture,covariates))

# Find the 25th and 75th quantiles of furan 1
furan_qntls <- quantile(mixture[,"Furan1"], probs=c(0.25,0.75))

# create a new dataset with two rows 
# each row has the means for all variables except furan 1
# each row has one of the furan 1 quantiles
new_data <- rbind(data_means,data_means)
new_data[,"Furan1"] <- furan_qntls

# check data
round(new_data,2)
```

Second, we predict with this new data.

```{r predict outcomes at each level}
posterior_furan1 <- predict(fit_bart, newdata = new_data)
dim(posterior_furan1)
colnames(posterior_furan1) <- c("q25","q75")
matplot(posterior_furan1, type="l")
```

Third, we take the difference at each iteration. Each row in the posterior is one iteration.
```{r take difference between levels at each iteration}
posterior_furan1_difference <- posterior_furan1[,"q75"] - posterior_furan1[,"q25"]
plot(posterior_furan1_difference, type="l")
```


Last, we summarize the posterior distribution of the difference.
```{r summarize posterior of difference}
mean(posterior_furan1_difference)
quantile(posterior_furan1_difference, c(0.025,0.975))
mean(posterior_furan1_difference>0)
```

## Estimation a contrast in one exposure averaged over all other predictors



We can estimate an IQR change in furan 1 averaged over all other covaraites

The first step it to setup the data.This time we do it separately for each level of furan 1. 
We also use the full dataset for prediction not just the means.

```{r BART IQR change at averaged over other variables data prep}
# create two replicates of the data
data_q25 <- cbind(mixture,covariates)
data_q75 <- cbind(mixture,covariates)

# Find the 25th and 75th quantiles of furan 1
furan_qntls <- quantile(mixture[,"Furan1"], probs=c(0.25,0.75))

# for each dataset assign all individuals the specified level of furan 1
data_q25[,"Furan1"] <- furan_qntls[1]
data_q75[,"Furan1"] <- furan_qntls[2]

# check data
round(head(data_q25),2)
round(head(data_q75),2)
```

Second, we predict with each new dataset.

```{r predict outcomes for both new datasets}
posterior_furan1_q25 <- predict(fit_bart, newdata = data_q25)
posterior_furan1_q75 <- predict(fit_bart, newdata = data_q75)
dim(posterior_furan1_q25)
```

Third, we take the average of all 1003 observations for all datasets at each iteration. Each row in the posterior is one iteration.
```{r take the mean of each response at each iteration separately for the two datasets}
posterior_furan1_q25 <- rowMeans(posterior_furan1_q25)
posterior_furan1_q75 <- rowMeans(posterior_furan1_q75)
posterior_furan1_means <- cbind(posterior_furan1_q25,posterior_furan1_q75)
dim(posterior_furan1_means)
colnames(posterior_furan1_means) <- c("q25","q75")
matplot(posterior_furan1_means, type="l")
```



Then we take the difference at each iteration. Each row in the posterior is one iteration.
```{r take difference of the means at each level of furan 1}
posterior_furan1_means_difference <- posterior_furan1_means[,"q75"] - posterior_furan1_means[,"q25"]
plot(posterior_furan1_means_difference, type="l")
```


Last, we summarize the posterior distribution of the difference.
```{r summarize posterior of difference the same as before}
mean(posterior_furan1_means_difference)
quantile(posterior_furan1_means_difference, c(0.025,0.975))
mean(posterior_furan1_means_difference>0)
```




## Partial dependence functions

To get partial dependence functions we can replicate the above procedure for estimates averaged over the other variables.
This is done several times, For example, we may do this at every 5th quantile of a variable and then plot the results.
The code below does this over a grid of L values, which can be specified.
The partial dependence functions with BART include uncertainty.

```{r BART partial dependence for Furan 1, eval=FALSE}
# partial dependence of BART
funan1_pd <- partialdependence1(fit_bart,
                                data=cbind(mixture,covariates),
                                exposures = "Furan1",
                                L=50)
```

```{r BART partial dependence for Furan 1 visualization}
ggplot(funan1_pd, aes(x=x, y=mean, ymin=lower, ymax=upper)) + 
  geom_ribbon(fill="grey70") + 
  geom_line() + 
  theme_minimal() + 
  ggtitle("Funan 1 partial dependence with BART") + 
  xlab("Exposure (z-score of log-transformed exposure)") + 
  ylab("Estimate (mean response)") 
```

```{r BART partial dependence for all components, echo=FALSE, eval=FALSE}
all_pd <- partialdependence1(fit_bart,
                            data=cbind(mixture,covariates),
                            exposures = exposure_names,
                            L=50)
```

```{r BART partial dependence for all components visualization}
plt <- all_pd %>%
mutate(exposure = fct_recode(exposure, "PCB 74" = "PCB74",
                                "PCB 99" = "PCB99",
                                "PCB 118" = "PCB118",
                                "PCB 138" = "PCB138",
                                "PCB 153" = "PCB153",
                                "PCB 170" = "PCB170",
                                "PCB 180" = "PCB180",
                                "PCB 187" = "PCB187",
                                "PCB 194" = "PCB194",
                                "Dioxin 1\n1,2,3,6,7,8-hxcdd" = "Dioxin1",
                                "Dioxin 2\n1,2,3,4,6,7,8-hpcdd" = "Dioxin2",
                               "Dioxin 3\n1,2,3,4,6,7,8,9-ocdd" =  "Dioxin3",
                               "Furan 1\n2,3,4,7,8-pncdf" =  "Furan1",
                               "Furan 2\n1,2,3,4,7,8-hxcdf" =  "Furan2",
                               "Furan 3\n1,2,3,6,7,8-hxcdf" =  "Furan3",
                               "Furan 4\n1,2,3,4,6,7,8-hxcdf" =  "Furan4",
                               "PCB 169" =  "PCB169",
                                "PCB 126" = "PCB126")) %>% 
  ggplot(aes(x=x, y=mean, ymin=lower, ymax=upper)) +
  facet_wrap(~exposure) +
  geom_ribbon(fill="grey70") +
  geom_line() +
  theme_minimal() +
  ggtitle("Partial dependence with BART") +
  xlab("Exposure (z-score of log-transformed exposure)") +
  ylab("Estimate (mean response)")
plt
```

## 2-way partial dependence functions

One form of a 2-way partial dependence plot shows the partial dependence of one exposure at different percentiles of another exposure. In this case, the partial dependence plot for PCB169 is shown at five percentiles of Furan 1. There is no evidence of interaction so the lines are exactly on top of each other.

```{r two way partial dependence with BART, eval=FALSE}
pd_2way <- partialdependence2(fit_bart,
                                data=cbind(mixture,covariates),
                                var = "PCB169",
                              var2 = "Furan1",
                              qtls = c(0.1,0.25,0.5,0.75,0.9),
                                L=20)
```


```{r two way partial dependence with BART visualize}
pd_2way$qtl <- as.factor(pd_2way$qtl)
ggplot(pd_2way, aes(x=x, y=mean, 
                    color=qtl,
                    linetype=qtl)) + 
  geom_line() + 
  theme_minimal() + 
  ggtitle("PCB169 partial dependence by quantile of Furan 1 with BART") + 
  xlab("Exposure (z-score of log-transformed exposure)") + 
  ylab("Estimate (mean response)") 
```


## Total mixture effect

We can estimate the total mixture effect in the same way that is typical with BKMR. This analysis estimates the mean response when all covariates are set to a specific quantile. In this case each quantile from 0.2 to 0.8 in steps of 0.05. 


```{r BART total mixture effect, eval=FALSE}
totalmix <- totalmixtureeffect(fit_bart,
                               data=cbind(mixture,covariates),
                               exposures = exposure_names,
                               qtls = seq(0.2,0.8,0.05))
```

```{r BART total mixture effect visualization}
ggplot(totalmix, aes(x=quantile, y=mean, ymin=lower, ymax=upper)) + 
  geom_errorbar(width=0) + geom_point() + theme_minimal() +
  ggtitle("Total mixture effect with BART") + 
  xlab("Quantile") + 
  ylab("Estimate (mean response)")
```



## Subgroup specific effects

BART naturally accounts for heterogeneity. Because we included all exposures and covariates in the model there can be interactions both among exposures and among exposures and covaraites. There are two approaches to estimate subgroup specific effects

First, we can estimate the exposure-response relationship in different subgroups. Below we estimate the partial dependence function for Furan 1 among males and females. This is from the same model fit as before. We are simply using the posterior sample to estimate a different estimand.

The first step is to subset the data based on the subgroups of interest.

```{r subset the data to male and female}
combind_data <- cbind(mixture,covariates)
male_data_subset <- combind_data[which(combind_data[,"male"]==1),]
female_data_subset <- combind_data[which(combind_data[,"male"]==0),]

# dimension and distribution of male in combined data
dim(combind_data)
table(combind_data[,"male"])
# dimension and distribution of male in male only data
dim(male_data_subset)
table(male_data_subset[,"male"])
# dimension and distribution of male in female only data
dim(female_data_subset)
table(female_data_subset[,"male"])
```

Next, estimate the partial dependence function for each subgroup. Alternatively, estimate the total mixture effect or interactions. 

```{r estimate partial dependence for each subset, eval=FALSE}
# estimate among males with male subset of data
funan1_pd_male_subset <- partialdependence1(fit_bart,
                                 data=male_data_subset,  # subset of data provided
                                 exposures = "Furan1",
                                 L=50)

# estimate among females with female subset of data
funan1_pd_female_subset <- partialdependence1(fit_bart,
                                     data=female_data_subset,   # subset of data provided
                                     exposures = "Furan1",
                                     L=50)
```


Finally, we can visualize the difference.


```{r graph subset specific partial dependence functions}
funan1_pd_male_subset$subgroup <- "Male"
funan1_pd_female_subset$subgroup <- "Female"
funan1_pd_mf_subset <- rbind(funan1_pd_male_subset,funan1_pd_female_subset)
ggplot(funan1_pd_mf_subset, aes(x=x, y=mean, ymin=lower, ymax=upper)) + 
  geom_ribbon(fill="grey70") + 
  geom_line() + 
  theme_minimal() + 
  ggtitle("Funan 1 partial dependence with BART") + 
  xlab("Exposure (z-score of log-transformed exposure)") + 
  ylab("Estimate (mean response)")  +
  facet_grid(.~subgroup)
```


An alternative approach is to use the full dataset but set all individuals to male or female. This eliminates differences in the distribution of covariates between the two groups (e.g. differences in age, BMI, exposures, etc.).

First, create the combined datasets.

```{r create datasets where all individuals are assigned one level of the modifier}
combind_data_allmale <- combind_data
combind_data_allmale[,"male"] <- 1

combind_data_allfemale <- combind_data
combind_data_allfemale[,"male"] <- 0



# dimension and distribution of male in combined data
dim(combind_data)
table(combind_data[,"male"])
# dimension and distribution of male in male only data
dim(combind_data_allmale)
table(combind_data_allmale[,"male"])
# dimension and distribution of male in female only data
dim(combind_data_allfemale)
table(combind_data_allfemale[,"male"])
```


Next, estimate the partial dependence using each dataset.

```{r estimate partial dependence functions on each dataset, eval=FALSE}
funan1_pd_male_all <- partialdependence1(fit_bart,
                                     data=combind_data_allmale,
                                     exposures = "Furan1",
                                     L=50)

funan1_pd_female_all <- partialdependence1(fit_bart,
                                       data=combind_data_allfemale,
                                       exposures = "Furan1",
                                       L=50)
```


Finally, graph the results.


```{r graph subset specific partial dependence functions using alternative approach}
funan1_pd_male_all$subgroup <- "Male"
funan1_pd_female_all$subgroup <- "Female"
funan1_pd_mf_all <- rbind(funan1_pd_male_all,funan1_pd_female_all)
ggplot(funan1_pd_mf_all, aes(x=x, y=mean, ymin=lower, ymax=upper)) + 
  geom_ribbon(fill="grey70") + 
  geom_line() + 
  theme_minimal() + 
  ggtitle("Funan 1 partial dependence with BART") + 
  xlab("Exposure (z-score of log-transformed exposure)") + 
  ylab("Estimate (mean response)")  +
  facet_grid(.~subgroup)
```





```{r save results, include=FALSE, purl=FALSE}
# this is used to save results from the full run
save(fit_bart, funan1_pd, all_pd, totalmix, pd_2way, 
     funan1_pd_male_all, funan1_pd_female_all, funan1_pd_male_subset, funan1_pd_female_subset, 
     file="bart_fits_for_lab.rda")
```

